Leadership across cultures and contexts: an exploratory analysis

### 2020-05-02 leadership_across_cultures_contexts.Rmd
Edits to discussion.

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. The variables on leader qualities and functions were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence type, and sex of the leader.

## The group context of human leadership

The distribution of text records by group and subsistence type suggests leaders emerge within similar social domains among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype).

@@

@@

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social movements). cheng_dominance_2019 and cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status, especially across nonhuman species, among humans high social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community nor be a contending Big man, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."

>>>>

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. The variables on leader qualities and functions were *not* determined prior to the study but were instead identified and coded during multiple passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence type, and sex of the leader.

## Effects of culture and context on human leadership

The distribution of text records by group and subsistence type suggests leaders emerge within similar social domains among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype). Exploratory analyses identified both universal and variable dimensions of human leadership, across cultures and contexts.

@@

Exploratory modeling revealed variation in the vast majority (INCLUDE NO. or %) of our researcher coded variables (TABLEs) did not vary by group type or subsistence strategy (ajusting for effects of leader sex and continental region and the hierarchical structure of ethnographer within culture), suggesting that many of these dimensions of leadership are potnetially universal features of human social groups or at least not strongly impacted by the within or between-group measures investigated here.

@@

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status, especially across nonhuman species, among humans high social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community nor be a candidate for community leadership, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."
